Horizontal aggregation of spatially biased ecosystem condition indicators – a GIS based workflow

### 2024-02-29 manuscript/Manuscript.qmd

---
title: Horizontal aggregation of spatially biased ecosystem condition indicators – a GIS based workflow
subtitle: A Short Subtitle
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    orchid: https://orcid.org/0000-0002-9623-9491
    affiliations: 
        - id: nina
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Pb 5685 Torgarden
          city: Trondheim
          postal-code: 7485
    attributes:
        corresponding: true
  - name: Matthew Grainger
    email: matthew.grainger@nina.no
    orchid: https://orcid.org/0000-0001-8426-6495
    affiliations:
        - ref: nina
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - alien species
  - disturbance
  - indicator
  - ecosystem condition
  - wetlands
  - mire
  - peatlands
  - ecosystem accounting
  - SEEA EA
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  include: false
  eval: false
  warning: false
  message: false
---

@@

# Take muni3 and add all sorts of other data to it
# using left_join.
muni_tbl <- muni3 |>
  # calculate total area
  mutate(
    area_km =
      round(
        units::drop_units(
          geom |> st_area() * 1e-6
        )
      )
  ) |>
  # make tibble for the coming join
  as_tibble() |>
  # paste inn terrestrial area
  left_join(
    terrestrial |>
      as_tibble() |>
      select(kommunenummer, t_area_km),
    keep = F
  ) |>
  # area of survey
  left_join(dk2 |> select(Municipality, dk_area_km)) |>
  mutate(dk_percent = round((dk_area_km / t_area_km) * 100)) |>
  # number of polygons
  left_join(nature3_tbl |> select(Municipality, n)) |>
  # total mire area and %
  left_join(mireArea |> select(Municipality, mirePercent, mire_km2)) |>
  # % mire inside survey coverage map
  left_join(mire_in_dk |> select(Municipality, mireInSurvey_km2)) |>
  mutate(mireInSurvery_percent = round(mireInSurvey_km2 / mire_km2 * 100, 2)) |>
  left_join(infraMuni3_summary)
```

@@

#ggsave("../images/forest-plot.jpg",
#       plot=forest_plot)
```

```{r forest_plotExample}
#| eval: true
#| cache:true
# Here is the same plot but only for trenching

@@

```{r EEA-plot2}
#| eval:true
#| cache: true
# Same as above, but for trenching only

@@

In this study we explore the potential for using a stratified aggregation technique to make use of spatially biased field data in ecosystem condition accounting.
We demonstrate this technique using a generic GIS-based workflow for compiling ecosystem condition accounts that can be applied at any spatial scale, and we highlight the opportunities for local use-cases of this workflow by contrasting our findings across three neighboring municipalities in Norway (@fig-studyLocation).
The main question is how much generalisation can we perform on the data we have before the resulting indicator looses its practical value in local governance processes.
We end by interviewing end-users from the relevant municipalities about the perceived benefits and shortcomings of our condition indicators.

# Material and Methods

This study makes use of a data set from from a standardised field survey of nature types in Norway that started in 2018 and which is still ongoing [@norwegian_environmental_agency_naturtyper_2024].
In this survey, selected nature types are delineated on a map (over 140 000 polygons at the end of 2023), and each locality is scored on a range of variables relevant for describing the state and quality of nature (REF INSTRUKS).
The surveys are commissioned with the goal of producing data relevant for immediate land-use decisions, and is therefore spatially biased, typically towards areas with high human impact or expected impact.
In addition there is a thematic and size bias in the sampling protocol.
For examples, for the forest ecosystem, rare, endangered or calcareous forest types are delineated, whereas more common or ordinary forest types are not.
For that reason we we focused on open mire ecosystems in Norway where the thematic bias is less severe.
Of all possible mire types, the survey only maps the following:

@@

In the above, *southern* refers to boreonemoral and southboreal zones, and *northern* refers to mid-boreal, north boreal, and alpine zones @asbjørnmoen1998.
In addition, the northern fens need to be even more calcareous than the southern fens in order to be surveyed.
We included data from 2018 to 2023.

In this paper we assume the survey is representative for the entire mire ecosystem in Norway.
Although it is possible that smaller or less calcareous mires will score systematically different than the ones that are surveyed we do not think this is so much the case for our variables.
However, we do assume that alien plants are slightly less common on bogs relative to fens, and therefore that this variable will biased since that variable for only recorded in fens.

From the survey data set we identified six relevnt variables (variable 1-6 in @tbl-variables) which we attribute to three different ecosystem condition characteristics that describe the typical behavior of open mire ecosystems in the reference condition @fig-workflow.

Variables 1 -5 were were originally recorded along binned frequency ranges.
because the data was strongly right skewed we used the lower limit for each frequency range to convert them into percentages.
Variables 2-5 describe very related aspects, and are attributed to the same ecosystem condition characteristic (vegetation intactness) and so they where combined into a single indicator called antropogenic disturbance to soil and vegetation, or ADSV for short.
This was done by summing the two variables after they had been converted to percentages.
This was not a perfect solution, especially since some localities only had one of the variables recorded, but we chose this, rather than for example using a *worst rule* principle, in order to better separate the localities in terms of their indicator values.

To turn the variables into indicators we scaled them using three normative reference values each: an upper (best possible condition), a lower (worst possible condition), and a threshold value that defines the breaking point between good and poor condition.
The reference values are numerical representation of the reference condition.
We define the reference condition as one where ecosystems are subject to little or no human influence, with a climate as in the period 1961-1990 and a native species pool similar as today.

For the indicators Alien species and ADSV, the lower and upper reference values were defined as 100% and 0%, respectively.
The threshold for *good ecosystem condition* was defined as 10%, which was then mapped to the value 0.6 on the rescaled indicators [@fig-scaling].
The variable 7GR-GI was rescaled into the indicator named *trenchin* by using the lower and upper reference values 1 and 5, respectively, and a threshold value of 2.5.
A variable value of 1 indicates an intact mire, and a value of 5 indicates a mire transitioning away from a wetland.
A value of 2 indicates observable change within the range expected for the same mapping unit, and a value of 3 indicates a mire transitioning into a neighboring (ecologically speaking) mapping unit.
See @fig-workflow for a schematic workflow for the indicator development.

*Alien species* was assigned to ECT class B1 - Compositional state characteristics and *ADSV* and *Trenching* were both assigned to ECT class A1 - Physical state characteristics [@czucz_common_2021].

We used an ecosystem delineation map for open mires in southern Norway, produced using remotely sensed data and a deep learning model [@bakkestuen_delineation_2023].
This model, which has a 90.9% precision when tested against independent field data, estimates 12.7% of the area in southern Norway is mire [@bakkestuen_delineation_2023].
Mires are ecologically and socially important in Norway simply doe to its large extent, and due to its role in climate mitigation as mires store a large amount of carbon (REF?).
There has not been a national assessment of the ecosystem condition of mires in Norway, but the authors recent contributed on a report which presented several new indicators that can be used in future assessments for this ecosystem [@nybo_indikatorer_2023; see also @kolstad2023]).
The current study builds on the work in that report.

```{r fig-studyLocation2}
#| eval: true
#| include: true
#| out.width: '80%'
#| fig.cap:
#| - "Map of the three focal municiaplities. For each municipality the map shows 
#| ocean in blue and non-ocean
#| in green. The survey coverage maps are in grey, and the mapped mires are
#| in red, with polygon borderes made extra thich to make them visible,
#| but then also exagerating their size."

@@

| id  | Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                                                                       | Reference                          |
|------------|------------|------------|------------|------------|------------|
| 1   | 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019                 |
| 2   | 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration                                                             | @runehalvorsen2019                 |
| 3   | PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7SE, but recorded at a higher resolution                                                                                                                  | @miljødirektoratet                 |
| 4   | 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks                                                    | @rune_halvorsen_dokumentasjon_2019 |
| 5   | PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7TK, but recorded at a higher resolution                                                                                                                  | @miljødirektoratet                 |
| 6   | 7GR-GI               | Ditching intensity                               | Unitless, ordinal scale from 1 to 5                                                     | Describes the effect that drainage ditches is estimate to have on the species composition and environmental variables ones the system reached its new equilibrium | @miljødirektoratet                 |
| 7   | Infrastructure Index | Infrastructure Index                             | Unitless linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                                                                        | @erikstad_index_2023               |
| 8   | HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                                                                          | this paper                         |

@@

```{r fig-workflow}
#| eval: true
#| include: true
#| out.width: '90%'
#| fig.cap:
#| - "Schematic workflow followed in this paper. Sub-index A is calculated using 
#| spatially explicit aggregation."

knitr::include_graphics("../images/workflow.jpg")
```

We chose three municipalities in Norway to test out the indicators.
THe municipalities differ in several aspects, such as the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-munis, @fig-studyLocation2, Figure S1).

@@

Because the nature type survey data is spatially biased, we cannot simply assume that they are area representative.
In an attempt to overcome this issue we divided the area of Norway into four non-overlapping Homogeneous Impact Areas (HIAs) based on an infrastructure index [@tbl-variables, @erikstad_index_2023].
This index measures the frequency of different infrastructure types inside 500 m radius circles around each 10×10 m pixel.
We categorised this continuous variable into four classes (0–3) using the following value ranges: 0 = \<1; 1 ≤ 1 ≥ 6; 6 ≤ 2 ≥12; 3 ≥ 12.
We plotted this categorical map over an area in Norway that we are familiar with and confirmed a sensible seperation of areas from areas with no roads of major infrastructure (HIA = 0) to urban areas (HIA = 3).
We then aggregated the data to 1×1 km pixels to ease computations.

We then took the nature type polygons with the indicator values and intersected with the HIA map and a map of municipal outlines in Norway.
For each HIA and municipality combination we then created a probability distribution for the indicator values using Bayesian updating of a uniform prior between 0 and 1.
The resulting distribution will be

@@

```{r fig-workflowExample}
#| eval: true
#| include: true
#| out.width: '100%'
#| fig.cap: |
#| - "Simplified schematic showing the workflow (Figures a-E) for horizontal aggregation of a spatially biased ecosystem condition indicator. Figure A shows a spatially explicit indicator. The colors indicator different indicator values (highlighted in the inset). The outline is the ecosystem accounting area (EEA). Figure B shows the distributio of four homogeneous impact areas (HIAs) inside the EEA. The indicator values in Figure A are used to update a uniform prior to produce a posterior probability distributio for the mean indicator value for each HIA (Figure C). The bayesian updating is informed by the standard deviation for the indicator in the much bigger national data set. Unlike an arithmetic approach where the variance in the mean indicator value per HIA is unreligable for very small sample sizes, the posterior distribution gives a religable measure for the indicator uncertainty even with a singe observation. The distributions in Figure C are assumed normally distributed. In Figure D the mean and SD from the posterior distributions are transferred to individual polygons in an ecosystem delineation map, for each HIA seperately. For each ecosystem occurence (i.e. ecosystem assets), we draw one random value for each square meter from a normal distribution with the mean and SD that is associated with that polygon. In Figure E we randomly sample 1000 values from the entire vector of sample in the previous step, and we get a area weighted probability distribution for the mean indicator value for the EEA."

@@

```{r fig-validation}
#| eval: true
#| include: true
#| out.width: '90%'
#| fig.cap:
#| - "Proportion of localities with different scaled indicator values for three 
#| ecosystem condition indicators along a gradient of infrastructure
#| densities."

@@

```{r fig-forest}
#| eval: true
#| include: true
#| out.width: '80%'
#| fig.cap:
#| - "Indicator values (circles = mean; bars = 95% credible intervals) for three
#| mire ecosystem condition indicators in three municipalities. The indicator values 
#| are calculated uniquely for each Homogeneous Impact Area (HIA) in each 
#| municipality. The numbers to the left of each bar is the sample size, i.e. 
#| the number of surveyed mires."

@@

```{r fig-spread-example}
#| eval: true
#| include: true
#| out.width: '80%'
#| fig.cap:
#| - "An indicator for mire trenching shown for Nord-Aurdal municipality. 
#| Individual mire polygons are colored by the mean indicator values for the
#| homogenous impact area where it lies. Colors are chosen to best reflect 
#| categorical differences and exagerates the absolute difference between 
#| areas. The inset it just a visual aid."

@@

```{r fig-EEA-ridgePlot}
#| eval: true
#| include: true
#| out.width: '80%'
#| fig.cap:
#| - "Distributions for three ecosystem condition indicators in the Norwegian 
#| Municiaplities. The color gradient reflects the same as the x-axis. 
#| The dotted vertical line represnet the threshold for what is considered 
#| reduced ecosystem condtion (< 0.6). "

@@

# Credit authorship contribution statement {.unnumbered}

@@

@@

{{< pagebreak >}} \# References {.unnumbered}

>>>>

---
title: Horizontal aggregation of spatially biased ecosystem condition indicators – a GIS based workflow
subtitle: A Short Subtitle
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    orchid: https://orcid.org/0000-0002-9623-9491
    affiliations: 
        - id: nina
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Pb 5685 Torgarden
          city: Trondheim
          postal-code: 7485
    attributes:
        corresponding: true
  - name: Matthew Grainger
    email: matthew.grainger@nina.no
    orchid: https://orcid.org/0000-0001-8426-6495
    affiliations:
        - ref: nina
  - name: Marianne Evju
    email: marianne.evju@nina.no
    orchid: https://orcid.org/0000-0001-7338-5376
    affiliations:
        - id: nina2
          name: Norwegian Institute for Nature Research
          department: NINA Oslo
          address: Sognsveien 68
          city: Oslo
          postal-code: NO-0855
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - alien species
  - disturbance
  - indicator
  - ecosystem condition
  - wetlands
  - mire
  - peatlands
  - ecosystem accounting
  - SEEA EA
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  include: false
  eval: false
  warning: false
  message: false
---

@@

# Take muni3 and add all sorts of other data to it
# using left_join.
muni_tbl <- muni3 |>
  # calculate total area
  mutate(
    area_km =
      round(
        units::drop_units(
          geom |> st_area() * 1e-6
        )
      )
  ) |>
  # make tibble for the coming join
  as_tibble() |>
  # paste inn terrestrial area
  left_join(
    terrestrial |>
      as_tibble() |>
      select(kommunenummer, t_area_km),
    keep = F
  ) |>
  # area of survey
  left_join(dk2 |> select(Municipality, dk_area_km)) |>
  mutate(dk_percent = round((dk_area_km / t_area_km) * 100)) |>
  # number of polygons
  left_join(nature3_tbl |> select(Municipality, n)) |>
  # total mire area and %
  left_join(mireArea |> select(Municipality, mirePercent, mire_km2)) |>
  # % mire inside survey coverage map
  left_join(mire_in_dk |> select(Municipality, mireInSurvey_km2)) |>
  mutate(mireInSurvery_percent = round(mireInSurvey_km2 / mire_km2 * 100, 1)) |>
  left_join(infraMuni3_summary) |>
  mutate(mire_km2 = round(mire_km2, 1),
         meanHIA = round(meanHIA, 1))
```

@@

ggsave("../images/forest-plot.jpg",
       plot=forest_plot,
       width = 8,
       height = 6)
```

```{r forest_plotExample}
#| eval: true
#| cache: true

# Here is the same plot but only for trenching

@@

```{r EEA-plot2}
#| eval: true
#| cache: true

# Same as above, but for trenching only

@@

In this study we explore the potential for using a stratified aggregation technique to make use of spatially biased field data in ecosystem condition accounting.
We demonstrate this technique using a generic GIS-based workflow for compiling ecosystem condition accounts that can be applied at any spatial scale, and we highlight the opportunities for local use-cases of this workflow by contrasting our findings across three neighboring municipalities in Norway.
The main question is how much generalisation can we perform on the data we have before the resulting indicator looses its practical value in local governance processes.
We end by interviewing end-users from the relevant municipalities about the perceived benefits and shortcomings of our condition indicators.

# Material and Methods {#sec-methods}

This study makes use of a data set from from a standardised field survey of nature types in Norway that started in 2018 and which is still ongoing [@norwegian_environmental_agency_naturtyper_2024].
In this survey, selected nature types are delineated on a map (over 140 000 polygons at the end of 2023), and each locality is scored on a range of variables relevant for describing the state and quality of nature [@norwegian_environmental_agency_protocol].
The surveys are commissioned with the goal of producing data relevant for immediate land-use decisions, and is therefore spatially biased, typically towards areas with high human impact or expected impact.
In addition there is a thematic and size bias in the sampling protocol.
For examples, for the forest ecosystem, rare, endangered or calcareous forest types are delineated, whereas more common or ordinary forest types are not.
For that reason we we focused on open mire ecosystems in Norway where the thematic bias is less severe.
Of all possible mire types, the survey only maps the following:

@@

In the above, *southern* refers to boreonemoral and southboreal zones, and *northern* refers to mid-boreal, north boreal, and alpine zones [@asbjørnmoen1998].
In addition, the northern fens need to be even more calcareous than the southern fens in order to be surveyed.
We included data from 2018 to 2023.
In this paper we assume the survey is representative for the entire mire ecosystem in Norway.
Although it is possible that smaller or less calcareous mires will score systematically different than the ones that are surveyed we do not think this is so much the case for our variables (see below).
However, we do assume that alien plants are slightly less common on bogs relative to fens, and therefore that this variable will be biased since that variable for only recorded in fens.
The other variables are recorded in all delineated mires.

From the survey data set we identified six relevant variables (variable 1-6 in @tbl-variables) which we attribute to three different ecosystem condition characteristics that describe the typical behavior of open mire ecosystems in the reference condition (Figure S1).
Variables 1 -5 were were originally recorded along binned frequency ranges @norwegian_environmental_agency_protocol .
Because the data was strongly right skewed we used the lower limit for each frequency range to convert them into percentages.
Variables 2-5 describe very related aspects, and are attributed to the same ecosystem condition characteristic (vegetation intactness) and so they where combined into a single indicator called antropogenic disturbance to soil and vegetation (ADSV; Figure S1).
This was done by summing the two variables after they had been converted to percentages.
This was not a perfect solution, especially since some localities only had one of the variables recorded, but we chose this, rather than for example using a *worst rule* principle, in order to better separate the localities in terms of their indicator values.
Variable 6 is different from the preceding variables in @tbl-variables in that it includes an estimation of future effects that the observed trenches are projected to have on mire vegetation, function or structure over time.
This is not a favorable trait in a metric used to evaluate the ecosystem condition as it is today, yet we include it here nonetheless because there is a general shortage of data on mire hydrology, which is a fundamental part of mire integrity.

To turn the remaining three variables into indicators we scaled them using three normative reference values each: an upper (best possible condition), a lower (worst possible condition), and a threshold value that defines the breaking point between good and poor condition.
The reference values are numerical representation of the reference condition.
We define the reference condition as one where ecosystems are subject to little or no human influence, with a climate as in the period 1961-1990 and a native species pool similar as today.

For the indicators Alien species and ADSV, the lower and upper reference values were defined as 100% and 0%, respectively.
The threshold for *good ecosystem condition* was defined as 10%, which was then mapped to the value 0.6 on the rescaled indicators, thus creating a non-linear rescaling of the variable (Figure S2).
The variable 7GR-GI was rescaled into the indicator named *Trenching* by using the lower and upper reference values 1 and 5, respectively, and a threshold value of 2.5 (Figure S2).
A variable value of 1 indicates an intact mire, and a value of 5 indicates a mire transitioning away from a wetland.
A value of 2 indicates observable change within the range expected for the same mapping unit, and a value of 3 indicates a mire transitioning into a neighboring (ecologically speaking) mapping unit.
See Figure S1 and @fig-workflowExample for a schematic workflow for the indicator development.

According to the SEEA ECT, *Alien species* can be attributed to ECT class B1 - Compositional state characteristics and *ADSV* and *Trenching* attributed to ECT class A1 - Physical state characteristics [@czucz_common_2021].

We used an ecosystem delineation map for open mires in southern Norway, produced using remotely sensed data and a deep learning model [@bakkestuen_delineation_2023].
This model, which has a 90.9% precision when tested against independent field data, estimates 12.7% of the area in southern Norway is mire [@bakkestuen_delineation_2023].
Mires are ecologically and socially important in Norway simply doe to its large extent, and due to its role in climate mitigation as mires store a large amount of carbon.
There has not been a national assessment of the ecosystem condition of mires in Norway, but the authors recent contributed on a report which presented several new indicators that can be used in future national assessments for this ecosystem [@nybo_indikatorer_2023; see also @kolstad2023]).
The current study builds on the work in that report.

We chose three municipalities in Norway to test out the indicators.
The municipalities differ in several aspects, such as the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-munis, @fig-studyLocation2, Figures S3, S4).

```{r fig-studyLocation2}
#| eval: true
#| include: true
#| out.width: '80%'
#| fig.cap:
#| - "Map of the three focal municiaplities. For each municipality the maps
#| in the top row shows ocean in blue and non-ocean in green. 
#| The survey coverage maps are in grey, and the mapped mires are
#| in red, with polygon borderes made extra thich to make them visible,
#| but then also exagerating their size.
#| The bottom row shows the delineation for homogeneous impact areas which
#| is an ordinal gradient from 0–3 with increasing presence of human 
#| instrastructure."

@@

| id  | Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                                                                       | Reference            |
|--------|------------|------------|------------|-----------------|----------------|
| 1   | 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019   |
| 2   | 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration                                                             | @runehalvorsen2019   |
| 3   | PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7SE, but recorded at a higher resolution                                                                                                                  | @miljødirektoratet   |
| 4   | 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks                                                    | @runehalvorsen2019   |
| 5   | PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7TK, but recorded at a higher resolution                                                                                                                  | @miljødirektoratet   |
| 6   | 7GR-GI               | Trenching intensity                              | Unitless, ordinal scale from 1 to 5                                                     | Describes the effect that drainage ditches is estimate to have on the species composition and environmental variables ones the system reached its new equilibrium | @miljødirektoratet   |
| 7   | Infrastructure Index | Infrastructure Index                             | Unitless linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                                                                        | @erikstad_index_2023 |
| 8   | HIA                  | Homogeneous Impact Area                          | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                                                                          | this paper           |

@@

@@

Because the nature type survey data is spatially biased, we cannot assume that they are area representative.
In an attempt to overcome this issue we divided the area of Norway into four non-overlapping Homogeneous Impact Areas (HIAs) based on an infrastructure index for the year 2022 [@tbl-variables, @erikstad_index_2023].
This index is a continuous variable that represnets the frequency of different infrastructure types inside 500 m radius circles around each 100×100 m pixel.
We then categorised this continuous variable into four classes (0–3) using the following value ranges: 0 = \<1; 1 ≤ 1 \> 6; 6 ≤ 2 \>12; 3 ≥ 12.
The name HIAs is not a perfect representation of the information found in the infrastructure index, but we introduce this name here as a general term which is aligned with the concept of Homogeneous Ecosystem Areas, *sensu* @vallecillo_eu-wide_2022.
We plotted the HIA map over an area in Norway that we are familiar with and confirmed a sensible separation of areas from areas with no roads of major infrastructure (HIA = 0) to urban areas (HIA = 3).
We then aggregated the data to 1×1 km pixels to ease computations, and vectorized it.

We then took the nature type polygons with the indicator values and intersected with the HIA map and a map of municipal outlines in Norway.
The relationship between the HIA classes and indicator values was examined visually.
For each HIA and municipality combination we then created a probability distribution for the area weighted mean indicator values using Bayesian updating of a uniform prior between 0 and 1, informed by the standard deviation (SD) of the indicator values in the national data set [@fig-workflowExample C; Appendix C, line XXX].
The resulting distributions are assumed normally distributed, and we therefore simply carried the mean and SD from the posterior distributions over to individual polygons in the ecosystem delineation map, for each HIA class separately, and for each polygon we then sampled random numbers from a normal distribution with this same mean and SD.
The number of m^3^ dictated the number of samples for each polygon, thus ensuring that large polygons end up counting more towards the indicator value in the entire municipality.
We then drew 1000 random values from this much large vector of possible indicator values and created a probability distribution for the indicators for the entire municipalities (i.e. the ecosystem accounting areas).
When there were no indicator values for a given HIA class, we ignored that class also in the municipal estimate.

@@

The three indicators showed some association with the HIAs when we lookad at data from all of Norway with `r nrow(naturetypes)` individual mires.
The indicators mostly showed a worsening of condition with increasing presence of human infrastructure (@fig-validation).
This relationship was much stringer for the indicator Trenching, and weakest for the indicator Alien species.
Alien species also had the highest indicator values overall, with most mirer having no alien species recorded.
For Trenching, 57% of mires in HIA-3 had some trenches, whereas in HIA-0 this number was 10%.
However, only in Gran municipality was there a significant difference in the Trenching indicator between HIAs, with HIA-2 having worse condition than HIA-1 (@fig-forest).
Conversely, the ADSV indicator in Gran showed worse condition in HIA-1 compared to HIA-2.
Nordre-Follo had considerably fewer data points compared to Gran, and especially to Nord-Aurdal who had the most data points, and this paucity of data is reflected in the wide credible intervals for all three indicators in Nordre Follo.
The credible intervals are widest for the Trenching indicators, and this is because the national data set showed more variation in the underlying variable 7GR-GI, and this information is informing the Bayesian updating process (see @sec-methods).
None of the three municipalities had data for all four HIAs.
Therefore, when we transferred the mean and SD for each HIA over to the ecosystem delineation map (@fig-spread-example), some mire polygons did not get assigned any data, and were therefore dropped from any further analyses.
At the municipality level (i.e. the ecosystem accounting area level) the three municipalities show some differentiation.
Nordre-Follo had the highest (the best) indicator values for ADSV, but the lowest for Trenching, which was also the only instance of an indicator crossing the threshold from good to deteriorated condition (@fig-EEA-ridgePlot; Table S1).

```{r fig-workflowExample}
#| eval: true
#| include: true
#| out.width: '100%'
#| fig.cap:
#| - "Simplified schematic showing the workflow (Figures A-E) for horizontal aggregation of a spatially biased ecosystem condition indicator. Figure A shows a spatially explicit indicator. The colors indicator different indicator values (highlighted in the inset). The outline is the ecosystem accounting area (EEA). Figure B shows the location of four homogeneous impact areas (HIAs) inside the EEA. The indicator values in Figure A are used, in combination with the HIA map in Figure B, to update a uniform prior and produce a posterior probability distribution for the mean indicator value for each HIA (Figure C; here simplified to only show the mean (circles) and 95% credible intervals (colored bands)). The bayesian updating is informed by the standard deviation for the indicator in a much bigger national data set. Unlike an arithmetic approach, the width of the posterior distribution responds both to variation in the data and to the sample size, giving  a religable measure for the uncertainty around the indicator even with a singe observation. The distributions in Figure C are assumed normally distributed. Note that because in this example there were no indicator data for HIA-3, indicator values are only aggregated for HIAs 0–2. In Figure D the mean and SD from the posterior distributions are transferred to individual polygons in an ecosystem delineation map, for each HIA seperately. For each ecosystem occurence (i.e. ecosystem assets) in Figure D, we draw one random value for each square meter of area from a normal distribution with the mean and SD that is associated with that polygon. In Figure E we randomly sample 1000 values from the entire vector of samples in the previous step, and we get an area weighted probability distribution for the mean indicator value for the EEA."

@@

```{r fig-validation}
#| eval: true
#| include: true
#| out.width: '90%'
#| fig.cap:
#| - "Proportion of localities with different indicator values for three 
#| ecosystem condition indicators along a gradient of infrastructure
#| densities."

@@

```{r fig-forest}
#| eval: true
#| include: true
#| out.width: '80%'
#| fig.cap:
#| - "Indicator values (circles = mean; bars = 95% credible intervals) for three
#| mire ecosystem condition indicators in three Norwegian municipalities. The indicator values 
#| are calculated uniquely for each Homogeneous Impact Area (HIA) in each 
#| municipality. The numbers to the left of each bar is the sample size, i.e. 
#| the number of surveyed mires."

@@

```{r fig-spread-example}
#| eval: true
#| include: true
#| out.width: '80%'
#| fig.cap:
#| - "An indicator for mire trenching shown for Nord-Aurdal municipality. 
#| Individual mire polygons are colored by the mean indicator values for the
#| homogenous impact area where it lies. Colors are chosen to best reflect 
#| categorical differences and exagerates the absolute difference between 
#| areas. The inset it just a visual aid. Coordinate reference system is 
#| EPSG 25832. Axis are in meters."

@@

```{r fig-EEA-ridgePlot}
#| eval: true
#| include: true
#| out.width: '80%'
#| fig.cap:
#| - "Distributions for three ecosystem condition indicators in the Norwegian 
#| municiaplities. The color gradient reflects the value of the x-axis. 
#| The dotted vertical line represnet the threshold for what is considered 
#| reduced ecosystem condtion (< 0.6). Vertical lines under the density curves
#| are 2.5%, 50% (the median) and 97.5% percentiles."

@@

# CRediT authorship contribution statement {.unnumbered}

**Anders L. Kolstad**: Conceptualization, Methodology, Software, Validation, Formal analysis, Data Curation, Writing - Original Draft, Visualization, Project administration, Funding acquisition.
**Matthew Grainger**: ...
**Marianne Evju**: ...

@@

Kwaku Peprah Adjei Hanno Sandvik

@@

This manuscript is written in Quarto, and the source file (Appendix B) also contains code for all the analyses underlying this study, including data exploration and cleaning.
For a rendered version of the source file, with all code and calculation visible, see Appendix C. The source files and the data are also located on GitHub (https://github.com/anders-kolstad/HIAs).
The exception is the large nature type survey data which was downloaded locally, but which is freely available (see @norwegian_environmental_agency_naturtyper_2024).

# References {.unnumbered}

::: {#refs}
:::
